---
title: "Case study 1: How Does a Bike-Share Navigate Speedy Success"
author: "Soa-Yu Chan"
date: "2021/6/19"
output: pdf_document
#output: 
#  html_document: 
#    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal: How do annual members and casual riders use Cyclistic bikes differently?
  
  
### Contents of Document

1. Data Source
2. Data Structure and manipulate
3. Data visualizations
4. Preliminary results
  
  
### 1. Data Source

This is a public data came from [**Google Data Analytics**](https://www.coursera.org/professional-certificates/google-data-analytics) course in coursera.  Cyclistic is a fictional bike-share company located in Chicago with more than 5,800 bicycles and 600 docking stations. The data has been made available by Motivate
International Inc. under this [**license**](https://www.divvybikes.com/data-license-agreement). This case study hope to find the difference between annual members and casual riders.
  
  
### 2. Data Structure and Manipulate

* `X.U.FEFF.`: Index  
* `ride_id`: ID attached to each trip taken  
* `rideable_type`: rideable type  
* `start_at`: day and time trip started, in CST  
* `ended_at`: day and time trip ended, in CST  
* `start_station_name`: name of station where trip originated  
* `start_station_id`: ID of station where trip originated  
* `end_station_name`: name of station where trip terminated  
* `end_station_id`: ID of station where trip terminated  
* `start_lat`: station latitude where trip originated  
* `start_lng`: station longitude where trip originated  
* `end_lat`: station latitude where trip terminated  
* `end_lng`: station longitude where trip terminated  
* `member_casual`: "casual" is a rider who purchased single-ride passes or full-day passes; "member" is a rider who purchased an Annual Membership  
* `ride_length_second`: each ride time from trip originated to trip terminated, in second  
* `ride_length`: each ride time from start to end, in hh:mm:ss
* `day_of_week`: the day of the week that each ride started  

From **2020-04** to **2021-05**. 

```{r import, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
library(table1)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)

bike_share <- read.csv(file = "D:/case_study_2021_06_13/202004_202105_divvy_tripdata.csv",
               header = T, na.strings = c("", "NA"), encoding = "UTF-8", sep = ",")
glimpse(bike_share)
```
***  
* Add `ride_length_hour`: ride length, in hour
* Add `started_date`: started date, in yyyy-mm-dd  
* Add `start_mm_yyyy`: started year-month, in yyyy-mm  

```{r mutate, message=FALSE, warning=FALSE, paged.print=TRUE}
bike_share$ride_length_hour <- ((bike_share$ride_length_second + 0.001)/3600) - (0.001 / 3600)
bike_share$started_date <- as_date(ymd_hms(bike_share$started_at))
bike_share$start_mm_yyyy <- format_ISO8601(bike_share$started_date, precision = "ym")
```
***  
* Some trip characteristic between casual riders and annual members

```{r summary, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
table1(~ factor(day_of_week) + ride_length_hour + factor(rideable_type) + start_mm_yyyy | factor(member_casual), data = bike_share)
```
***  
* Casual Riders Top 10 Trip  

```{r table1, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
bike_share %>% 
  group_by(start_station_name, end_station_name) %>%
  filter(member_casual == 'casual') %>% 
  drop_na() %>% 
  summarize(count_start_end = n(), average_rider_length = mean(ride_length_hour) * 60) %>% 
  arrange(desc(count_start_end)) %>% 
  `colnames<-`(c("Start station name", "End station name", "Count", "Average minutes per ride")) %>% 
  head(n=10)
```
  
* Annual Members Top 10 Trip  

```{r table2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
bike_share %>% 
  group_by(start_station_name, end_station_name) %>%
  filter(member_casual == 'member') %>% 
  drop_na() %>% 
  summarize(count_start_end = n(), average_rider_length = mean(ride_length_hour)*60) %>% 
  arrange(desc(count_start_end)) %>% 
  `colnames<-`(c("Start station name", "End station name", "Count", "Average minutes per ride")) %>% 
  head(n=10)
```
  
  
  
### 3. Data Visualizations (Bar Chart - Annual Member vs. Casual rider)
  
  
##### Visualization on Month Year  
* Casual rider has less trip during winter season compare to annual member(Fig.1).  
* Annual member spend less time than casual rider for each trip(Fig.2).  
```{r plt1, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot(data = bike_share)+
  geom_bar( position = 'dodge', mapping = aes(x = factor(start_mm_yyyy), fill = member_casual)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title = 'Fig.1: Total number of trips according to Month',
       x = 'Year-Month', y = 'Number of trips', fill='Rider Type') +
  scale_y_continuous(breaks = c(0,100000,200000,300000,400000),
                     labels = c("0","100K","200K","300K","400K")) + 
  theme(legend.position = c(.95, .97),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.margin = margin(2, 10, 2, 6))

t2 <- bike_share %>% 
  group_by(start_mm_yyyy, member_casual) %>% 
  summarize(average_ride_length = mean(ride_length_hour*60), sum_ride_length = sum(ride_length_hour)) 

ggplot(t2, aes(x = start_mm_yyyy, y = average_ride_length,  fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(), width =  0.5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title = 'Fig.2: Average Minutes per Trip by Year-Month',
       x = 'Year-Month', y = 'Average minutes per trip', fill = 'Rider Type') +
  theme(legend.position = c(.95, .97),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.margin = margin(2, 10, 2, 6))

```
  
***  
  
##### Visualization on Day of Week  
* Much more trip during weekend for casual riders(Fig.3).  
* No significant difference from sunday to saturday for annual members(Fig.3).  
* Also, casual rider spend more time than annual member for each trip(Fig.4).  
  
```{r plt2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
ggplot(data = bike_share) +
  geom_bar(mapping = aes(x = factor(day_of_week), fill = member_casual)) +
  facet_wrap(~member_casual) +
  labs(title = 'Fig.3: Total number of trips according to The Day of Week',
   x = 'The Day of Week', y = 'Number of trips', fill = 'Rider Type') +
  scale_x_discrete(breaks = 1:7,
                   labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000),
                   labels = c("0", "100K", "200K", "300K", "400K")) + 
  theme(legend.position = 'none')

t3 <- bike_share %>% 
  group_by(day_of_week, member_casual) %>% 
  summarize(average_ride_length = mean(ride_length_hour*60), sum_ride_length = sum(ride_length_hour)) 

ggplot(t3, aes(x = factor(day_of_week), y = average_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.5) +
  facet_wrap(~member_casual) +
  labs(title = 'Fig.4: Average Minutes per Trip by The Day of Week',
       x = 'The Day of Week', y = 'Average minutes per trip', fill = 'Rider Type') +
  scale_x_discrete(breaks = 1:7,
                   labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) +
  theme(legend.position = 'none')

```
  
  
***  
  
##### Visualization on Rideable Type  
* For annual members: No docked_bike trip after 2021-01. Only classic_bike and electric_bike(Fig.5).  
* For casual members: Total number of Classic_bike trip portion larger than docked_bike trip since 2020-12(Fig.5).  
* docked_bike spend more average time than classic_bike per trip(Fig.6).  
```{r plt5, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
t5 <- bike_share %>% 
  group_by(start_mm_yyyy, member_casual, rideable_type) %>% 
  summarize(average_ride_length = mean(ride_length_hour * 60), sum_ride_length = sum(ride_length_hour), sum_trip = n()) 

ggplot(t5, aes(x = start_mm_yyyy, y = sum_trip,  fill = rideable_type)) +
  geom_bar(stat = "identity", position = position_dodge(), width =  0.5) +
  facet_wrap(~member_casual) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title = 'Fig.5: Total Number of Trips by Year-Month for different Rideable Type',
       x = 'Year-Month', y = 'Total number of trips', fill = 'Rideable Type') +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000),
                     labels = c("0", "100K", "200K", "300K", "400K")) +
  theme(legend.position = c(.95, .97),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.margin = margin(2, 10, 2, 6))

ggplot(t5, aes(x = start_mm_yyyy, y = average_ride_length,  fill = rideable_type)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.5) +
  facet_wrap(~member_casual) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title = 'Fig.6: Average Ride Minutes per Trip by Year-Month for different Rideable Type',
       x = 'Year-Month', y = 'Average ride minutes per trip', fill = 'Rideable Type') +
  theme(legend.position = c(.95, .97),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.margin = margin(2, 10, 2, 6))
```
  
***  
  
### 4. Preliminary results  
* Change all docked_bike to classic_bike. It can save time for riders.  
* Add a half-year annual member to pricing plans. Because many casual riders not use in winter.  
* Set more docking station near office zone. Annual member use Cyclistic to commute to
work each day.  